---
title: "R Exercise 1: INTA 6450 - Data Analytics and Security"
output: pdf_document
bibliography: references/references.bib
csl: references/ieee.csl
urlcolor: blue
date: "2024-10-15"
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

# R Exercise 1

### Prompt

1. Download `exercise_1.r`
2. Open the file in RStudio and run the commands in the exercise and look at 
the output.
3. Read the comments to try to understand how each command is working, and 
look at the output. 
Make some changes to the code to explore how the output differs.
4. Then, submit the output for running these commands for this assignment. 
You may copy/paste output from your terminal into a word or text document. 
You can convert the notebook to PDF submission:File->Compile Report->Report 
output format as HTML/PDF. 
5. Write and submit 100-150 words describing the extension you made to the code.
You may post the change description as a comment at the beginning of your code,
or submit the write-up in a separate document. Make sure to clearly mark where 
in your code the changes were made either by commenting, highlighting, or noting
the command lines where the changes occur. 
**NOTE:** Exported code does not retain line numbers. 
Please refrain from phrases such as "code change begins on line 62".

### Submissions

You will submit the following:

1. Original code output (R file code and output)
2. Your code change and output from that code change 
(comment where your code changes start and end)
3. Your code change description (this should be between 100 to 150 words)

An example of what the change description should look like:
```
#############################
# CHANGE DESCRIPTION
# 100-150 words on the changes I’ve made, including a reference to which 
# initial command I’ve changed >
# END OF CHANGE DESCRIPTION
#############################
```

Files must be in one of the following formats: pdf, doc, html

*Please note some of these exercises have graphs/charts - 
be sure when you export the original output to also check that the file 
includes graphs/charts that may have been present. 
If it does not, you will need to re-export or manually include the graphs in 
your final pdf, doc or html file*

\newpage
# Solution

### Wages

To observe the relationship between `wage` and each of the 17 predictors, 
a matrix scatter plots is developed with `ggplot` and `gridExtra`.
A function is created to create a single scatter plot so that mapping can be
utilized to create all 17 plots at once and be placed in a matrix for a single
image.

- Change `lm` model to 17 predictors
- Change `lm` model to significant number of predictors
- Add cv
- `lm` with step wise regression w/ 12 predictors

### Labor Force Participation

- `lm` w/ 7 predictors
- `glm` w/ 7 predictors
- `glm` using `probit` model w/ 7 predictors


### Wages Overview

The data obtained from `wage2.csv` contains information on monthly earnings,
education, several demographic variables, and IQ scores for 
935 men in 1980 [@blackburn1992].

[@woolridge]

- `wage`
- `hours`
- `IQ`
- `KWW`
- `educ`
- `exper`
- `tenure`
- `age`
- `married`
- `black`
- `south`
- `urban`
- `sibs`
- `brthord`
- `meduc`
- `feduc`
- `lwage`

Predicted wage based on simple linear regression model using `lm` in the `stats`
package. 
Using `wage` as the response variable and `educ` or education as the
predictor variable.
Second try is using `wage` as the response variable and both `educ` and `IQ` as
predictors.
Final approach was to use step wise regression with `lm` model.

- `lm` regression model
- Use `ggplot2` to evaluate the relationship between `wage` and all predictors.
  - See HW5 ISYE6501 Question 8.2
- Model data is adequate as we typically look for 10:1 ratio on data points vs. factors
- cv for `lm` uses `DAAG` and `glm` uses `boot`



### Load and Inspect the Data

```{r, message=FALSE}
# Import libraries
# lm and glm are in the stats package loaded by default
library(ggplot2) # Plot functions
library(reshape)

# Wages ------------------------------------------------------------------------
# Load wages data from AWS .csv
wages <- read.csv('http://inta.gatech.s3.amazonaws.com/wage2.csv')
# Write the wages data to a .csv in local directory
# write.csv(wages, "exercises/M5_regression/data/wage2.csv", row.names = FALSE)

# Summary of wages statistics from data frame
summary(wages)
```

### Form Linear Regression Model with All Predictors

```{r, eval = FALSE, message=FALSE, warning=FALSE}
# Predict observed crime rate
# wage~. separates wage (response variable) from predictors
model <- lm(wage~., data=wages)
summary(model)
```

### Observing the Relationship between Wage and Predictors

To observe the relationship between `wage` and each of the 17 predictors, 
a matrix scatter plots is developed with `ggplot` and `gridExtra`.
Remember, `ggplot2` was imported earlier, so two additional libraries need to be
loaded. The library `gridExtra` is to arrange a grid, and `purrr` is used to 
map the plots.
First, a list of the predictor variables is defined as `predictors`.
The list of predictors is then utilized to generate scatter plots.
A function is created to create a single scatter plot so that mapping can be
utilized to create all 17 plots at once and be placed in a matrix for a single
image.
```{r, eval = FALSE, message=FALSE, warning=FALSE}
library(gridExtra)
library(purrr)

# Create a list of predictor variables
predictors <- c("hours", "IQ", "KWW", "educ", "exper", "tenure", "age",
                "married", "black", "south", "urban", "sibs", "brthord",
                "meduc", "feduc", "lwage")

# Function to create a scatter plot with regression line for each predictor
plot_predictor <- function(predictor) {
  ggplot(wages, aes_string(x = predictor, y = "wage")) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    theme_minimal() +
    labs(title = paste("wage vs.", predictor))
}

# Create plots for all predictors
plots <- map(predictors, plot_predictor)

# Arrange plots in a grid
scatter <- grid.arrange(grobs = plots, ncol = 4)
scatter
```
```{r echo = FALSE, warning = FALSE, message = FALSE, results = 'hide', preview = TRUE, fig.show="hold", out.width="100%", fig.align="center"}
library(gridExtra)
library(purrr)

# Create a list of predictor variables (17)
predictors <- c("hours", "IQ", "KWW", "educ", "exper", "tenure", "age",
                "married", "black", "south", "urban", "sibs", "brthord",
                "meduc", "feduc", "lwage")

# Function to create a scatter plot with regression line for each predictor
plot_predictor <- function(predictor) {
  ggplot(wages, aes_string(x = predictor, y = "wage")) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    theme_minimal() +
    labs(title = paste("wage vs.", predictor)) +
    theme(plot.title = element_text(size = 10))
}

# Create plots for all predictors
plots <- map(predictors, plot_predictor)

# Arrange plots in a grid
scatter <- grid.arrange(grobs = plots, ncol = 4)
scatter
```



# References
