---
title: "R Exercise 2: INTA 6450 - Data Analytics and Security"
output: pdf_document
bibliography: references/references.bib
csl: references/ieee.csl
urlcolor: blue
# date: "2024-10-02"
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

# R Exercise 2

### Prompt

1. Complete the following exercise: 
  - `exercise_2.r`
2. You may make changes to the code to explore how they work
3. Then, submit the output for running these commands. 
You may copy/paste output from your terminal into a word or text document. 
You can convert the notebook to PDF submission:File->Compile Report->Report 
output format as HTML/PDF. 
4. Write 100-150 words describing one extension you made to the code. 
You may post the change description as a comment at the beginning of your code, 
or submit the write-up in a separate document. Make sure to clearly mark 
where in your code the changes were made either by commenting, highlighting, 
or noting the command lines where the changes occur. 

### Submissions

You will submit the following:

1. Original code output (R file code and output)
2. Your code change and output from that code change 
(comment where your code changes start and end)
3. Your code change description (this should be between 100 to 150 words)

An example of what the change description should look like:
```
#############################
# CHANGE DESCRIPTION
# 100-150 words on the changes I’ve made, including a reference to which 
# initial command I’ve changed >
# END OF CHANGE DESCRIPTION
#############################
```

Files must be in one of the following formats: pdf, doc, html

*Please note some of these exercises have graphs/charts - 
be sure when you export the original output to also check that the file 
includes graphs/charts that may have been present. 
If it does not, you will need to re-export or manually include the graphs in 
your final pdf, doc or html file*

\newpage
# Solution Summary

The solution first inspected residuals to check the model fit.
The purpose was to understand the residuals or differences between
observed and predicted values to assess the fit for both OLS and IV models.
The Distributions are inspected which follow a normal distribution with
most values centered around zero. This indicated a good fit.

Second, the results for `ols` with a linear regression model is evaluated.
By using `ols.robust` we can see that education, marital status, age and
experience are all positive impact on wages. They are also statistically
significant. In addition, `black` has a significant effect on wages.
The small standard errors for most coefficients indicate that the estimates are 
adequate and precise, and the high t-values and small p-values confirm the 
statistical significance of these relationships. The overall summary is that 
education and experience are particularly 
important drivers of wages, with a negative impact associated with race (Black) 
and a positive impact from marital status.

**Original code is attached in submission for reference or to run, as it is
fully functional.**

### Original Working Code

```{r, eval = TRUE, message=FALSE, warning=FALSE}
# Instrumental Variables example
library(sem)
#If gets error, add code< install.packages("Hmisc",dependecies=T)> or 
# download a new version R studio
wages<-read.csv('http://inta.gatech.s3.amazonaws.com/wage2.csv')
iv.results<-tsls(lwage ~ educ + age + married + black + exper, ~ feduc + age +
                   married + black + exper, data=wages)
# Run an instrumental variables regression. Note that father's education is
# used for an instrument, and is not in the first set of variables. Age,
# marital, status, and race are assumed to be less related to underlying
# ability and so are controlled for in the "first stage" regression too.

ols.results<-lm(lwage ~ educ + age + married + black + exper, data=wages)
# Run the analogous OLS regression without father's education

print(summary(ols.results))
print(summary(iv.results))
# Note that the point estimate for education has now gone up. This suggests
# that people who have higher education in a way that's correlated with
# their father having higher education, even holding fixed age, marital
# status, and race, receive ~13% higher wages for every year of education.
# Also note that the standard error on education is higher for the iv than
# ols results. The reason for this is that in IV, we're using only some of
# the variation in education: only the part related to father's education,
# so it's like there's less variation overall

# Stepwise regression
library(MASS)
start.model<-lm(wage ~ hours + IQ + KWW + educ + exper + tenure + age + married + black + south + urban + sibs, data=wages)
# Give an initial model, which will be the most coefficients we'd want to ever use
summary(start.model)
stepwise.model<- step(start.model)
# The command "step" adds and subtracts coefficients to maximize a measure of
# goodness of fit
summary(stepwise.model)

stepwise.model.interactions <- step(start.model, scope=wage~.^2)
# The command "step" adds and subtracts coefficients to maximize a measure of
# goodness of fit. 
summary(stepwise.model.interactions)
```

### Inspect Residuals

```{r, eval = TRUE, message=FALSE, warning=FALSE}
# Residuals for OLS model
ols.residuals <- residuals(ols.results)
hist(ols.residuals, main = "Histogram of OLS Residuals", 
     xlab = "Residuals", col = "blue")

# Residuals for IV model
iv.residuals <- residuals(iv.results)
hist(iv.residuals, main = "Histogram of IV Residuals", 
     xlab = "Residuals", col = "red")
```

### Evaluating Results

```{r, eval = TRUE, message=FALSE, warning=FALSE}
# Load sandwich package
library(sandwich)
library(lmtest) # coeftest

# Robust standard errors for OLS
ols.robust <- coeftest(ols.results, vcov = vcovHC(ols.results, type = "HC1"))
print(ols.robust)
```

